---
title: "RMySQL in RStudio"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(Lahman)
```

In the previous post I demonstrated how to install and use `MySQL` from the command line. In this tutorial I will show you how to set up and query a relational database from RStudio.

## The database

These queries are done using the Lahman baseball open data set available [here](https://github.com/chadwickbureau/baseballdatabank). 

I chose these data because 1) they are open to anyone and can be run on your local machine, not behind a vpn, and 2) it is available as .csv files for loading into any statistical software. 

## Connecting to databases with RStudio

For this tutorial I will be using a local instance of MySQL through the `DBI`, `dplyr`, and `RMySQL` packages. A `DBI` connection is built below to my local instance of `MySQL`. 

```{r packages}
library(DBI)
library(RMySQL)
library(RMariaDB)
library(dplyr)
library(rstudioapi)
```


## Connect to `MySQL` with `DBI::dbConnect()`

The first connection will be done with the `DBI` package. 

```{r LahmanDBIMySQL}
LahmanDBIMySQL <- dbConnect(MySQL(), 
                    dbname = "lahman2016", 
                    host = "localhost", 
                    user = "root", 
                    # password = "%LWGma9c"
                    password = rstudioapi::askForPassword("Database password")
                    )
LahmanDBIMySQL
```

Create an object from `Lahman::Master`.

```{r Master}
Master <- Lahman::Master
Master %>% glimpse(78)
```

Create another object from `Lahman::Batting`.

```{r Batting}
Batting <- Lahman::Batting
Batting %>% dplyr::glimpse(78)
```

This `MySQL` statement creates a data frame of the Griffeys (father and son).

```sql
SELECT 
  -- Master columns
  mas.playerID, 
  mas.birthYear, 
  mas.nameFirst, 
  mas.nameLast, 
  mas.weight, 
  mas.height, 
  mas.bats, 
  mas.throws, 
  mas.retroID, 
  mas.bbrefID,
  -- Batting columns
  bat.teamID,
  bat.lgID,
  bat.yearID,
  bat.G,
  bat.AB,
  bat.R,
  bat.H,
  bat.HR,
  bat.RBI,
  bat.BB,
  bat.SO
  
FROM 
    Master AS mas
      INNER JOIN 
    Batting AS bat ON mas.playerID = bat.playerID
  WHERE nameLast = "Griffey";
```

And this creates the same table using `dplyr` commands. 

```{r Griffey-select}
Master %>% 
    dplyr::select(playerID, birthYear, nameFirst, nameLast,
                  weight, height, bats, throws, retroID, bbrefID) %>% 
    dplyr::filter(nameLast %in% "Griffey") %>% 
    dplyr::inner_join(x = ., y = Batting, by = "playerID") %>% 
    dplyr::select(teamID, lgID, yearID, G, AB, 
                  R, H, HR, RBI, BB, SO) %>% dplyr::glimpse(78)
```

