---
title: "RMySQL in RStudio"
output: github_document
author: "Martin Frigaard"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = FALSE,
                      size = "small",
                      fig.path = "images/")
library(tidyverse)
library(Lahman)
library(markdowntemplates)
```

In the previous post I demonstrated how to install and use `MySQL` from the command line. In this tutorial I will show you how to set up and query a relational database management system (RDBMS) from RStudio.

## The database

These queries are done using the Lahman baseball data set available [here](http://www.seanlahman.com/baseball-archive/statistics/). 

I chose these data because 1) they are open to anyone and can be run on your local machine, and 2) they are available as a .sql file and .csv files for loading into any RDBMS software. 

## Database packages in RStudio

For this tutorial I will be using a local instance of `MySQL` with help from the `DBI`, `dplyr`, and `RMySQL` packages.  

```{r packages}
library(DBI)
library(RMySQL)
library(RMariaDB)
library(dplyr)
library(rstudioapi)
library(ggthemes)
```


## Change MySQL settings 

Before we can connect to a database with RStudio, we need to change a few settings for MySQL in the **System Preferences**.

Select **MySQL** >> **Initialize Database** 

Now enter a new password for the **root** user and select **Use Legacy Password Encryption**. Then click **Start MySQL Sever**.

## Connect to MySQL with DBI::dbConnect()

A `DBI` connection is built below to my local instance of `MySQL`.

```{r LahmanDBIMySQL}
LahmanDBIMySQL <- DBI::dbConnect(MySQL(), 
                    dbname = "lahman2016", 
                    host = "localhost", 
                    user = "root", 
                  # rstudioapi::askForPassword(prompt = "database password"))
                    password = "Med!c012")
LahmanDBIMySQL
```

As I can see, this is a `<MySQLConnection:0,1>`. This is the object we will use to access the `lahman2016` database.

## Querying a database from RStudio

We can get a list of tables in the `lahman2016` database by using the `DBI::dbListTables()` function. 

```{r dbListTables}
DBI::dbListTables(LahmanDBIMySQL)
```

This is similar to the `SHOW TABLES;` command in `MySQL`. In the previous tutorial, I built a table of statistics on Ken Griffey Junior and Senior in the Lahman database. The `Griffeys` table contains following fields.

```{r dbListFields}
DBI::dbListFields(LahmanDBIMySQL, "Griffeys") 
```

Queries can be passed directly to the `DBI::dbGetQuery()` function, like the `SELECT` statement below.

```{r select-query}
DBI::dbGetQuery(LahmanDBIMySQL, "SELECT * FROM Griffeys;") %>%
  dplyr::glimpse(78)
```


I realize I don't have all the columns I want from the `Batting` and `Fielding` tables, so I will query `lahman2016` again to collect these data. 

## Aliasing columns in MySQL tables

There are three columns in the `Batting` table I want to add to `Griffeys`: `SF`, `2B`, and `3B`. Two of these names violate naming rules in R. 

```{r list-fields-Batting}
DBI::dbListFields(LahmanDBIMySQL, "Batting") 
```

The `2B` and `3B` columns contain the number of doubles and triples for each player, and `SF` is the number of sacrifice flies. Unfortunately, R won't allow columns to have names that start with a number. I can write a query to inlcude these columns, but use an alias to create R-friendly column names. 

```{r add-doubles-triples-Griffeys}
DBI::dbGetQuery(LahmanDBIMySQL, "SELECT 
  grf.playerID, 
  grf.birthYear, 
  grf.nameFirst, 
  grf.nameLast, 
  grf.weight, 
  grf.height, 
  grf.bats, 
  grf.throws, 
  grf.lgID,
  grf.yearID,
  grf.teamID,
  grf.G,
  grf.AB,
  grf.R,
  grf.H,
  grf.HR,
  grf.RBI,
  grf.BB,
  grf.SO,
  bat.2B AS dubs,
  bat.3B AS trips,
  bat.SF AS sac_flies
  FROM Griffeys AS grf
    INNER JOIN Batting AS bat
     ON grf.playerID = bat.playerID
     AND grf.yearID = bat.yearID
     AND grf.teamID = bat.teamID;") %>% glimpse(78)
```

## Storing SQL in character vectors

I can also assign the query above to a character vector (`batting_query`) and pass the vector to the `DBI::dbGetQuery()` function. I will assign these new variables to the `Griffeys` table.

```{r batting_query}
batting_query <- "SELECT 
  grf.playerID, 
  grf.birthYear, 
  grf.nameFirst, 
  grf.nameLast, 
  grf.weight, 
  grf.height, 
  grf.bats, 
  grf.throws, 
  grf.lgID,
  grf.yearID,
  grf.teamID,
  grf.G,
  grf.AB,
  grf.R,
  grf.H,
  grf.HR,
  grf.RBI,
  grf.BB,
  grf.SO,
  bat.2B AS dubs,
  bat.3B AS trips,
  bat.SF AS sac_flies
  FROM Griffeys AS grf
    INNER JOIN Batting AS bat
     ON grf.playerID = bat.playerID
     AND grf.yearID = bat.yearID
     AND grf.teamID = bat.teamID;"
Griffeys <- DBI::dbGetQuery(LahmanDBIMySQL, batting_query) 
Griffeys %>% glimpse(78)
```

Now that we see the SQL works and these two tables are joined, we can calculate a few new statistics in R.

## Calculating On-base plus slugging (OPS) 

A players [on-base plus slugging (OPS)](https://en.wikipedia.org/wiki/On-base_plus_slugging) is a measure of their overall batting performance. This is the sum of the player's on-base percentage and slugging average. The steps to calculate this statsitic are below. 

```{r GriffsOPSbyYear}
GriffsOPSbyYear <- Griffeys %>% 
  dplyr::mutate(
    # slugging percentage
    slug_perc = (H - dubs - trips - HR + 2 * dubs + 3 * trips + 4 * HR) / AB,
    # convert sac_flies to numeric
    sac_flies = as.numeric(sac_flies),
    # On Base Percentage
    ob_perc = (H + BB) / (H + AB + BB + sac_flies),
    # On-base plus slugging
    obslug_perc = slug_perc + ob_perc,
    # create a nameID
    nameID = 
      case_when(
        birthYear == 1950 ~ "Ken Griffey Sr",
        birthYear == 1969 ~ "Ken Griffey Jr"),
    # convert to a factor
    nameID = factor(nameID), 
    # make KGSr the reference
    nameID = relevel(nameID, ref = "Ken Griffey Sr"))
```

I created a few intermediate calculations before creating `obslug_perc`, and I also created a `nameID` so the two players can be easily graphed. Below is the on-base plus slugging (OPS) statistic for both Griffeys over their career. 

```{r plot-obs-by-year}
GriffsOPSbyYear %>% 
  ggplot2::ggplot(aes(x = yearID, y = obslug_perc, 
                      group = nameID, color = teamID)) +  
  geom_line() + 
  ggplot2::facet_grid(. ~ nameID, scales = "free") + 
  ggplot2::labs(
    title = "The Griffey's career on-base plus slugging (OPS)"
  ) + ggthemes::theme_fivethirtyeight()
```

## The Lahman package in R

There is also a `Lahman` package in R. I will use this below to demonstrate some of the similarities between `dplyr` and `MySQL` syntax. 

Create an object for the `Lahman::Master`, `Lahman::Fielding`, and `Lahman::Batting` tables. 

```{r Master-Batting-FieldingData}
Master <- Lahman::Master
Batting <- Lahman::Batting
Fielding <- Lahman::Fielding
```


The code below creates the same tables and graph using `dplyr` commands. 

```{r Griffey-GriffeyOPSPlot}
GriffeyOPSPlot <- Master %>% 
    dplyr::select(playerID, 
                  birthYear, 
                  nameLast) %>% 
  # find Griffeys
    dplyr::filter(nameLast %in% "Griffey") %>% 
  # join to batting table
    dplyr::inner_join(x = ., 
                      y = Batting, 
                      by = "playerID") %>% 
  # select/rename relevant columns
    dplyr::select(teamID,
                  nameLast,
                  playerID,
                  birthYear,
                  yearID,
                  G,
                  AB, 
                  R, 
                  H, 
                  HR, 
                  dubs = X2B, 
                  trips = X3B, 
                  RBI, 
                  BB, 
                  SO, 
                  sac_flies = SF) %>% 
    dplyr::mutate(
    # slugging percentage
    slug_perc = (H - dubs - trips - HR + 2 * dubs + 3 * trips + 4 * HR) / AB,
    # convert sac_flies to numeric
    sac_flies = as.numeric(sac_flies),
    # On Base Percentage
    ob_perc = (H + BB) / (H + AB + BB + sac_flies),
    # On-base plus slugging
    obslug_perc = slug_perc + ob_perc,
    # create a nameID
    nameID = 
      case_when(
        birthYear == 1950 ~ "Ken Griffey Sr",
        birthYear == 1969 ~ "Ken Griffey Jr"),
    # convert to a factor
    nameID = factor(nameID), 
    # make KGSr the reference
    nameID = relevel(nameID, ref = "Ken Griffey Sr")) %>% 
  # create plot of new OPS 
  ggplot2::ggplot(aes(x = yearID, y = obslug_perc, 
                      group = nameID, color = teamID)) +  
  geom_line() + 
  ggplot2::facet_grid(. ~ nameID, scales = "free") + 
  ggplot2::labs(
    title = "The Griffey's career on-base plus slugging (OPS)"
  ) + 
  ggthemes::theme_fivethirtyeight()
ggplot2::ggsave(filename = "GriffeyOPSPlot.png",  width = 7, height = 5, units = "in")
```

## End